---
title: "STAT350 Tutorial 1"
# author: "Dylan Maciel"
date: "11/09/2020"
linestretch: 1.5
urlcolor: blue
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this course we will be using R to implement the concepts learned in class.
This tutorial covers downloading and installing R and RStudio for Mac and 
Windows. Then we'll go over some basic R commands and do an example simple
linear regression on a data set.


# Installing R and RStudio


R is a free open-source software used for statistical computing; the current 
version is 4.0.2. You can code directly in the program, but using an IDE like
RStudio, which is also free, allows you to write more than one line at a time
and integrates many other useful tools that make using R more efficient. 

The order in which you install these programs is important. For Mac users you'll
have to download and install [XQuartz](https://www.xquartz.org/) first. Then, 
for both Windows and Mac install [R](https://mirror.rcg.sfu.ca/mirror/CRAN/) 
before you install [RStudio](https://rstudio.com/products/rstudio/download/).
For RStudio you'll want to download the free version of RStudio Desktop. I
recommend keeping all the defaults when installing these programs. If done in
the proper order, when you first open RStudio it will detect R on your computer
and start the R console itself.

For a more in-depth introduction to R and Rstudio I recommend reading through 
sections 1 to 4 of [A (very) short introduction to R](https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf) 
by Torfs and Brauer. It also goes through installation, but will also introduce 
you to some basic R functions and uses of RStudio.

If you have any troubles with these installations, I will be holding online 
office hours over zoom next week on Tuesday at 5pm and Wednesday at 10 am. I can 
also answer any other R questions you may have.


# An Initial Example

In what follows I will briefly go over reading in data and fitting a simple 
linear regression model. To get the most out of this example I would open a new
R script in RStudio and go through the code below. Copy the code, make sure you 
can get it working on your computer and try to understand what each line is 
doing.

## Data Entry

Typically, we use data from external files in R. The function to read in data 
from an ".csv" or excel file is \verb|read.csv|. This file can either be located 
on you local disk or we can read a file directly from the internet as seen below
. 

```{r}
forbes_data <- read.csv(file = "http://users.stat.umn.edu/~sandy/alr4ed/data/Forbes.csv",
                        header = TRUE,
                        sep = ',',
                        row.names = 1)
```

This function has many input. The first input \verb|file| tells R the name of 
file containing the dataset. If this file was on your computer the way to ensure 
that R can find the file is to put the whole path to the file within the quotes. 

We set the \verb|header| input to \verb|TRUE| if the first row of the dataset 
contains variable names, otherwise it should be set to \verb|FALSE|. If this is 
not explicitly set it will default to \verb|TRUE|.

The \verb|sep| input tells the function how the columns of data are separated
within the file. It's default value is \verb|','|.

The \verb|row.names| input names can be used to specify the column number that 
contains the row names. Many data sets do not have row names, for those this 
input does not need to be specified.

Here I have saved the data to the object named \verb|forbes_data|. Now anytime I 
want to work with this specific dataset I just need to call it by that name. The
name you choose for the dataset is important and should be informative.


## Looking at the data.

We can look at the first few rows of the data using the \verb|head()| function.

```{r} 
head(forbes_data)
```

So, we see that the dataset contains three variable; \verb|bp|, \verb|pres|, and
\verb|lpres|. Before fitting a model to the data it is good practice to 
do some initial data analysis. This is done to ensure the data looks the way we 
expect it to look. If the data has mistakes or missing values they could
negatively impact the performance of our model in the end. Datasets used in
tutorials will be clean and ready for analysis, but you should keep this in mind 
in your future work.

We can get univariate numerical summaries of the data using the \verb|summary()| 
function.

```{r}
summary(forbes_data)
```

We can also make graphical summaries. Below, I use the \verb|boxplot()| function 
to create a boxplot for each variable.

```{r}
boxplot(forbes_data,
        main = 'Boxplot of forbes_data')
```

```{r}
hist(forbes_data$lpres)
```


## Simple Linear Regression

For our example \verb|lpres| will be the response and \verb|bp| will be the 
predictor. First we'll create a basic scatter plot of the data using the
\verb|plot()| function.

```{r}
plot(formula = lpres ~ bp,
     data = forbes_data,
     main = "lpres vs. bp",
     xlab = "bp",
     ylab = "lpres")
```

The first input, \verb|formula|, tells R how to arrange the data on the plot. 
The response goes to the left of the \verb|~| and the predictor to the right.
The next argument tells R where the data is located. This is followed by 
arguments that let us set the main title, x-axis label, and y-axis label, 
respectively. There are other arguments to this function that let us further 
customize how we want the plot to look.

Looking at the plot, we see that a linear regression model would be suitable for
this data and can proceed with creating the model.

In R we use the function \verb|lm()| to fit linear regression models. The first 
two arguments for this function follow the same format that we used in the
\verb|plot()| function above.


```{r}
my_mdl <- lm(formula = lpres ~ bp, 
             data = forbes_data)
```


Here I have saved the resulting object from using this command to the name 
\verb|my_mdl|. If we want brief summary of the model we can call it by
name. 

```{r}
my_mdl
```

With this we see how the linear model was defined, but more importantly we get 
the  estimated coefficients of the model. Here the estimated intercept is
`r round(my_mdl$coefficients[1], 3)` and the estimated slope is 
`r round(my_mdl$coefficients[2], 3)`. If we want to extract the coefficient 
values from the model object we use \verb|$| followed by the name.

```{r}
my_mdl$coefficients
```

To get a single value, we then add the index within square brackets to the code
above. If we want the intercept, it's the first element and therefore indexed
by 1.


```{r}
my_mdl$coefficients[1]
```

We can get a more elaborate summary of the model we can use the \verb|summary()|
function on the model object. 

```{r}
summary(my_mdl)
```

The added information here will be useful in the coming weeks when we cover 
confidence intervals, hypothesis tests, and assessing the model.


Now that the model is fit we can add the regression line to the scatter plot of
the data. This is done with the function \verb|abline()| whose input \verb|reg|
is the model object defined above. Alternatively, we could also manually set the 
regression line values using the input \verb|a| for the intercept and \verb|b| 
for the slope. This function places a straight line over top of a scatter plot
and should therefore be called after the \verb|plot()| function. Here I've also 
included the argument \verb|col| to specify the colour of the line. 

```{r}
plot(formula = lpres ~ bp,
     data = forbes_data,
     main = 'lpres vs. bp',
     xlab = 'bp',
     ylab = 'lpres')

abline(reg = my_mdl, 
       col = 'red')
```



